PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2023-03-08 supplementary.tex
chore
Update on Overleaf.

\begin{figure}[htb]
    \includegraphics[width=0.5\textwidth]{figs_suppl/graph_representation.pdf}
    \caption{{\bf Pangenome graph representation}. Given a set of genomes \textit{PanGraph} will construct a pangenome graph by condensing all stretches of homologous sequences in \textit{pancontigs}, containing the alignment of these sequences. Genomes can then be represented as \textit{paths}, i.e. lists of pancontigs.
        In our representation a pangenome graph is in essence a collection of \textit{pancontigs} and \textit{paths}.
    }
    \label{fig:graph-representation}
\end{figure}

@@

\begin{figure}[htb]
    \includegraphics[width=0.5\textwidth]{figs_suppl/pancontig_alignment.pdf}
    \caption{{\bf From pancontig consensus to alignment}. In our pancontig representation, each sequence in the alignment is encoded as variations over the pancontig consensus. In practice in the pangraph output \texttt{json} file each alignment sequence is saved as a list of gaps to be added to the consenus, sites to be mutated, insertions to be performed in gaps, and deletions. By performing these changes in order on the consensus each sequence in the alignment can be retrieved. In this example we display how each of these variations are encoded and performed.
    }
    \label{fig:alignment-reconstruction}
\end{figure}

Each \textbf{path} is an oriented collection of pancontigs that represents a genome. For each item in the sequence along with the pacontig id PanGraph also stores the pancontig orientation (whether homology with the consensus is on the forward the reverse strain of the genome) and which variations on the pancontig consesus (i.e. which sequence in the pancontig alignment) correspond to this precise location in the genome.\\

From knowledge of all the pancontigs and pahts, every single genome can be exactly reconstructed. Every part of the sequence is always encoded in a pancontig. ``Small-scale'' difference between genomes (i.e. SNPs, small insertions and deletions) are encoded in pancontigs as variations over the consensus. ``Large-scale'' differences, such as presence or absence of long stretch of sequences, are instead encoded in paths as presence or absence of pancontigs. This scale separation can be tuned using PanGraph parameters, as will be explained in following sections.

@@

In this section we describe in more detail the algorithmic procedure used to build a pangenome graph. In essence, the full pangenome graph is built by placing genomes on terminal leaves of a guide tree, built from a sequence minimizer distance matrix. These genomes are initially considered as elementary graphs composed of a single \textit{pancontig}. These graphs are then propagated along the tree towards the root, and at each intersection a merge operation is performed. The full pangenome graph is then collected from the root.

We focus on three aspects:
\begin{itemize}
    \item The estimation of pairwise distances between genomes, used to build the guide tree.
    \item The procedure used to merge two pangenome graphs together, which is at the core of \textit{PanGraphs}'s algorithm.
    \item The effect of the main parameters of the algorithm on the resulting graph structure.
\end{itemize}

@@

At each merge operation two pangenome graphs need to be compressed into a single one. Three of pangraph's parameters play a role in the outcome of this procedure:
\begin{itemize}
    \item the minimal pancontig length $\Lthr$.
    \item pseudo-energy parameters $\alpha$ and $\beta$ (cf. eq.~(1)in the main text).
    \item the choice of alignment kernel (minimap2 or mmseqs) and aligner precision (asm5, asm10 or asm20 for minimap2).
\end{itemize}
In the merging, these parameters control how much of the diversity between sequence must be stored as large-scare variation (difference in paths) and how much as fine-grained variation (differences in alignments, stored in pancontigs). This is done for example by setting a divergence limit for sequences to be stored in the same alignment.\\
An implicit assumption of our procedure is that the phylogeny of sequences stored inside the same panconting is roughly star-like, with each sequence containing independent variations from the consensus. In such cases the consensus is a good ``average'' representation for the set of sequences in the alignment.
In the next section we briefly discussion how PanGraph behaves when this assumption is not satisfied, such as in the case of merging ortholog and paralog sequences in the same pancontig.\\
% move this paragraph or change?

@@

              \item Flanking regions of the consensus sequences that are not part of the matches are cut and not merged and stored in separate pancontigs. One exceptions is when any of these flanking regions is shorter than the threshold length $\Lthr$. In this case instead of creating a very short pancontig the region is stored as an indel in the merged pancontig alignment.
              \item The remaining homolous region is then a candidate for the merging. However the merging might result in more than a single panconting, depending on the underlying alignment. If the alignment contains insertions or deletions longer thant $\Lthr$, these need to appear as separate pancontigs. To this end the CIGAR string is parsed, and the long match is potentially split in multiple sub-matches, with interposed long insertions or deletions that will be saved as separate pancontigs.

@@

    \item Once the list of all matches has been exhausted the two graphs have been merged into one. However at the previous stage only one merge could be performed per pancontig, which could be insufficient to collapse all of the homology contained in the graph. To this end we repeat the procedure described in step 3, but instead of looking for matches between pancontigs of two different graphs, we search for matches between all pancontigs in the same graph. Merges are performed following the same rules, and this procedure is repeated until no new merges can be performed.

@@

As graphs are propagated along the guide tree more and more merges are performed. This makes pancontigs progressively shorter but their alignments deeper.\\
We also point out that in our merging procedure only the consensus sequence of any two pancontigs is used in their alignment. This makes this procedure fast and is sufficient to capture the large-scale structural variation of the pangenome, but might introduce minor inconsistencies and artifacts in the alignment. To this end \textit{PanGraph} has the command \verb|polish|\footnote{See \url{https://neherlab.github.io/pangraph/tutorials/tutorial_3/}.}. This command ``polishes'' alignments by solving on each pancontig the multiple sequence alignment problem using MAFFT \cite{katoh2002mafft}. It has a simple interface: it takes as input a pangenome graph and produces another pangenome graph with the same structure but improved pancontig alignments. This efficiently separates the problem of finding large-scale structural variation, and nucleotide-level variation. When precise knowledge of the latter is needed, a graph can be \textit{polished} restricting the multiple sequence alignment problem only on the homologous sequences contained in each pancontig.

@@

After explaining the procedure we use for merging, we conclude this section with some comments on the effect of PanGraph's parameters on the resulting pangenome graphs.\\

@@

The choice of alignment kernel and the value of $\beta$ set instead a limit on the divergence of sequences that are merged in the same pancontig. The first intrinsic limit is set by the choice of alignment kernel and its sensitivity: minimap2 is significantly faster than mmseqs2, but it is in general unable to find matches between homologous sequences with more than 20\% divergence (with asm20 sensitivity).\\
On top of this intrinsic limit of the aligner, a more fine-tuned setting of the acceptable amount of divergence can be achieved by tuning the value of $\beta$. From the definition of the pseudo-energy (cf. eq.~(1) in the main text) it follows that matches between homologous sequence with divergence higher than $1/\beta$ will have negative pseudo-energy and will thus not result in a merge. If the molecular clock hypothesis holds, tuning this parameter is equivalent to setting an effective time limit beyond which sequences are not to be merged in the same alignment. This parameter can be tuned based on the diversity of the sample in study. It can for example be useful in resolving situations in which the hypothesis of a star-like phylogeny in a pancontig alignment is broken, such as to separate loosely related paralogs.

@@

This paralog splitting can be done using two different sources of information: either sequence diversity or local context in which the duplication is inserted. Currently \textit{PanGraph} does not provide a command to perform paralog splitting. This is a feature that we plan on implementing in the future. However partial control of paralog split can currently be achieved in two ways.\\
The first one is based on sequence divergence. By tuning the value of the energy parameter $\beta$ the user can set a divergence limit. This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$ during graph construction.\\
The second method relies instead on context. The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag. When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context. These duplicated pancontigs are then split and absorbed in their context, thus simplifying the graph and loosing trace of the duplication.\\

@@

To test the accuracy of \textit{PanGraph} in reconstructing the pangenome graph of a set of isolates we generated artificial data using the procedure described in Section~\ref{sec:data_generation}. For each simulation we obtain both a set of genome sequences and the underlying real pangenome graph. This was done using standard value of the parameters, but varying the rate of HGT in the interval $h=[0.01,1]$ and the mutation rate $\mu=[0,0.01]$. For each $(h,\mu)$ pair 25 different sets of data were generated.\\
On each set of data we executed \textit{PanGraph} with values for the energy parameters $\alpha=0$ and $\beta=0$, and with three different alignment kernels:
\begin{itemize}
    \itemsep0pt
    \item \textit{minimap2} with option \textit{asm10},
    \item \textit{minimap2} with option \textit{asm20},
    \item \textit{mmseqs2},
\end{itemize}
thus obtaining three different pangenome graphs per dataset.\\
To link the mutation rate parameter $\mu$ with the average pairwise SNPs distance $\avg{d}$ between homologous segments in the population we evaluated the average pairwise distance for every pancontig in the pangenome graph with depth greater than one. For each graph we then perform the weighted average of these divergences, using as weight the length of the pancontig. Finally, for every pair of parameters $(\mu,h)$ we average these numbers over the 25 different simulations. Results are displayed in Fig.~\ref{fig:snps-suppl}.\\
For high values of $\mu$, the average pairwise divergence of pancontigs is not influenced by the rate of HGT $h$, but depends strongly on the choice of alignment kernel. With the \textit{asm20} option, \textit{minimap2} is able to correctly merge genomes with average pairwise divergence $\avg{d} \sim 7\%$ in our simulations, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher processing time. Notice that these values are lower than the threshold sensitivity of these aligners, which are expected to find matches up to respectively around 10\%, 20\% and 30\% sequence divergence. This is due to the fact that homologous sequences in our simulations are generated by an evolutionary process, and they can have a wide distribution of diversities and hierarchical population structure. Inability to merge sequences above a threshold divergence in a single panconting is expected to result in multiple pancontigs containing sub-clades with higher similarity, and average diversity that is appreciably lower than the aligner threshold.\\
By performing a linear fit on the datapoints with $\mu \leq 0.002$ we are able to recover the conversion factor between the mutation rate and average pairwise divergence of pancontigs in our simulations: $\avg{d} \sim 20.8 \, \mu$.

@@

\begin{figure}[htb]
    \includegraphics[width=\textwidth]{figs_suppl/accuracy_comparison.pdf}
    \caption{{\bf Accuracy of PanGraph algorithm on synthetic data}. For each of our generated datasets we evaluate the average displacement of inferred pancontig breakpoints. The figure represents the cumulative distribution of average misplacements, stratified by the average pairwise divergence of sequences in pancontigs (see colorbar on the right). Each panel corresponds to a different alignment kernel used to infer the pangenome graph (see panel title). As divergence increases we observe a transition in the precision of breakpoint position estimation. At low divergence the position of most breakpoints is correctly estimated withing a few bps, while at high divergence most breakpoints can be misplaced by multiple hundreds of bps. The transition value for the divergence depends on the alignment kernel used, with \textit{mmseqs2} having the highest accuracy at the cost of higher computational time.}
    \label{fig:benchmark-accuracy-suppl}
\end{figure}

@@

We downloaded sets of annotated and complete chromosomes from RefSeq \cite{o2016reference} for 5 different bacterial species: \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Prochlorococcus marinus} (PM), \textit{Mycobacterium tuberculosis} (MT) and \textit{Escherichia coli} (EC). In Table~I in the main text we report some summary statistics for these datasets. The number of isolates per species ranges from several hundreds for EC to 10 for PM.\\
These data has been chosen as part of a PanX \cite{ding2018panx} analysis dated March 2018.\footnote{
    We report for completeness that two of the isolates from the \textit{E.~coli} dataset from this analysis, namely \texttt{NC\_020518} and  \texttt{NC\_017663}, were subsequently removed from RefSeq. We verified that their removal from the dataset does not significantly impact core-genome and soft-core genome size, but decided to utilize the complete dataset as per original PanX analysis.
}
PanX is a pipeline for the analysis of bacterial pangenome. The pipeline takes as input a set of annotated genomes, and it clusters and aligns homologous genes. We used the PanX analysis to estimate the dataset diversity.\\
From this analysis we evaluated the average pairwise sequence divergence $\avg{d_\text{core}}$ of core genes, the average genome length $\avg{\Lgen}$, the total pangenome length $\Lpang$, the total core genome length $\Lcore$, and the total soft-core genome length $\Lsoftcore$, where soft-core genes are defined as genes being present in at least 95\% of the isolates. These quantities are reported in Table~I in the main text.\footnote{
    $\Lcore$ is the length of the alignment used to estimate sequence divergence $\avg{d_\text{core}}$. However this quantity is sensitive to annotation errors. This is particularly evident in the \textit{E.~coli} dataset, for which $\Lcore = 0.7$ Mbp is likelt to be a subestimation of the real core genome size. $\Lsoftcore$ is instead less sensitive to annotation errors and usually considered a better core-genome size estimate. For example the \textit{E.~coli} core genome is usually considered to be aroud 1500-2500 genes in size, depending on the definition used \cite{park2019large,sutton2021pan,horsfield2023accurate}.
}\\
EC is the dataset with the highest number of isolates (307) and is also the one with the biggest pangenome. MT is the species with the smallest sequence divergence ($\dcore \sim 0.03\%$) and the biggest core genome fraction. The two datasets with highest sequence divergence are HP ($\dcore \sim 4\%$) and PM ($\dcore \sim 24\%$). For HP the average sequence divergence sits on the limit of what \textit{PanGraph} is able to accurately merge when using the \textit{minimap2} kernel (cf. Fig.~\ref{fig:benchmark-accuracy-suppl}), but is merged correctly by the \textit{mmseqs2} kernel. The sequence divergence of PM sits instead beyond the capabilities of all alignment kernels.

@@

We used genomes from the dataset described in the previous section to verify that marginalized graphs are compatible with graphs built directly from subset of strains. For each species we selected 50 different isolates (10 for PM) and built pangenome graphs using standard parameter values and the \textit{minimap2} kernel with \textit{asm20} option. For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{The list of selected accession numbers and selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}} For each pair we then built pairwise graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf. Fig.~6A, main text).\\
The comparison was performed by considering that each graph partitions a genome into pancontigs. Each pancontig can either be shared with the other member of the pair, or can be private to the isolate (cf. Fig.~6A, main text). For each isolate we consider the two partitions defined by the marginalized and pairwise graphs, and generate their intersection. Each block in this intersection can belong to two different categories, depending on whether the two partitions agree or disagree on whether the block is shared. Moreover, we can separate blocks on which we find agreement in two further categories: blocks that are shared and blocks that are private on both graphs.\\
For each pair of genomes and each type of block we evaluate the fraction of the genome that they cover and their average size. In Fig.~\ref{fig:marginalize-suppl} we display the distribution of these quantities across all pairs for the different species considered and the block types described above.\\
In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default pancontig length threshold of \textit{PanGraph} ($\Lthr =$100 bp).
As described in the previous section, the divergence of the HP dataset sits on the limit of what the \textit{minimap2} alignment kernel can successfully handle. In this case the results of merging different graphs together can depend on the order of merging, and pairwise and marginalized graphs can potentially show some disagreement. In this case the disagreement remains minor and only extends to few percents of the genome size, with disagreement blocks having size only slightly bigger than the 100 bp threshold of \textit{PanGraph}.
PM represents instead an example of a dataset with divergence much beyond the limit of the alignment kernel ($\dcore \sim 27 \%$). As a consequence, highly diverged homologous sequences are not merged, and the great majority of segments are private and have large size. Results also show a strong order-dependence, with some \textit{disagreement} blocks having size of 10s of kbp.

>>>>

\begin{figure}[htb]
    \includegraphics[width=0.5\textwidth]{figs_suppl/graph_representation.pdf}
    \caption{{\bf Pangenome graph representation}. Given a set of genomes \textit{PanGraph} will construct a pangenome graph by condensing all homologous sequences in \textit{pancontigs}, containing the alignment of these sequences. Genomes can then be represented as \textit{paths}, i.e. lists of pancontigs.
        In our representation a pangenome graph is in essence a collection of \textit{pancontigs} and \textit{paths}.
    }
    \label{fig:graph-representation}
\end{figure}

@@

\begin{figure}[htb]
    \includegraphics[width=0.5\textwidth]{figs_suppl/pancontig_alignment.pdf}
    \caption{{\bf From pancontig consensus to alignment}. In our pancontig representation, each sequence in the alignment is encoded as variations over the pancontig consensus. In practice in the pangraph output \texttt{json} file each alignment sequence is saved as a list of gaps to be added to the consensus, sites to be mutated, insertions to be performed in gaps, and deletions. By performing these changes in order on the consensus each sequence in the alignment can be reconstructed. In this example we display how each of these variations are encoded and performed.
    }
    \label{fig:alignment-reconstruction}
\end{figure}

Each \textbf{path} is an oriented collection of pancontigs that represents a genome. For each item in the sequence along with the pancontig id, PanGraph also stores the pancontig orientation (whether homology with the consensus is on the forward the reverse strand of the genome) and which variations on the pancontig consensus (i.e. which sequence in the pancontig alignment) are present in this precise location in the genome of the isolate.\\

From knowledge of all the pancontigs and paths, every single genome can be exactly reconstructed. Every part of the sequence is always encoded in a pancontig. ``Small-scale'' difference between genomes (i.e. SNPs, small insertions and deletions) are encoded in pancontigs as variations over the consensus. ``Large-scale'' differences, such as presence or absence of long stretch of sequences, are instead encoded in paths as presence or absence of pancontigs. This scale separation can be tuned using PanGraph parameters, as will be explained in following sections.

@@

In this section we describe in more detail the algorithmic procedure used to build a pangenome graph. In essence, the full pangenome graph is built by placing genomes on terminal leaves of a guide tree, built from a sequence minimizer distance matrix. These genomes are initially considered as elementary graphs composed of a single \textit{pancontig}. These are then propagated along the tree towards the root, and at each intersection a merge operation is performed. The full pangenome graph is then collected from the root.\\

We focus on four aspects:
\begin{itemize}
    \item The estimation of pairwise distances between genomes, used to build the guide tree.
    \item The procedure used to merge two pangenome graphs together, which is at the core of \textit{PanGraphs}'s algorithm.
    \item The effect of the main parameters of the algorithm on the resulting graph structure.
    \item How pangraph processes duplicated regions and possible strategies for paralog splitting.
\end{itemize}

@@

At each merge operation two pangenome graphs need to be joined into a single one. Three of \textit{PanGraph}'s parameters play a role in the outcome of this procedure:
\begin{itemize}
    \item the minimal pancontig length $\Lthr$.
    \item pseudo-energy parameters $\alpha$ and $\beta$ (cf. eq.~(1) in the main text).
    \item the choice of alignment kernel (minimap2 or mmseqs) and aligner precision (asm5, asm10 or asm20 for minimap2).
\end{itemize}
In the merging, these parameters control how much of the diversity between sequences must be stored as ``large-scale'' variation (difference in paths) and how much as ``fine-grained'' variation (differences in alignments, stored in pancontigs). This is done for example by setting a divergence limit for sequences to be stored in the same alignment.\\
An implicit assumption of our procedure is that the phylogeny of sequences stored inside the same pancontig is roughly star-like, with each sequence containing independent variations from the consensus. In such cases the consensus is a good ``average'' representation for the set of sequences in the alignment.
In the next section we briefly discussion how PanGraph behaves when this assumption is not satisfied, such as in the case of merging paralogs in the same pancontig.\\

@@

              \item Flanking regions of the consensus sequences that are not part of the match are cut and not merged and stored in separate pancontigs (see fig.~1 right panel in the main text). One exceptions is when any of these flanking regions is shorter than the threshold length $\Lthr$. In this case instead of creating a very short pancontig the region is stored as an indel in the merged pancontig alignment.
              \item The remaining homologous region is then a candidate for the merging. However the merging might result in more than a single pancontig, depending on the underlying alignment. If the alignment contains insertions or deletions longer that $\Lthr$, these need to appear as separate pancontigs. To this end the CIGAR string is parsed, and the long match is potentially split in multiple sub-matches, with interposed long insertions or deletions that will be saved as separate pancontigs.

@@

    \item Once the list of all matches has been exhausted the two graphs have been merged into one. However at the previous stage only one merge could be performed per pancontig, which could be insufficient to collapse all of the homology contained in the graph. To this end we repeat the procedure described in step 3, but instead of looking for matches between pancontigs of two different graphs, we perform an all-against-all homology search between all pancontigs in the same graph. Merges are performed following the same rules, and this procedure is repeated until no new merges can be performed.

@@

As graphs are propagated along the guide tree more and more merges are performed. This makes pancontigs progressively shorter but their alignments deeper.\\

We point out that in our merging procedure only the consensus sequence of any two pancontigs is used in their alignment. This makes this procedure fast and is sufficient to capture the large-scale structural variation of the pangenome, but might introduce minor inconsistencies and artifacts in the alignment. To this end \textit{PanGraph} has the command \verb|polish|\footnote{See \url{https://neherlab.github.io/pangraph/tutorials/tutorial_3/}.}. This command ``polishes'' alignments by solving on each pancontig the multiple sequence alignment problem using MAFFT \cite{katoh2002mafft}. It has a simple interface: it takes as input a pangenome graph and produces another pangenome graph with the same structure but improved pancontig alignments. This efficiently separates the problem of finding large-scale structural variation, and nucleotide-level variation. When precise knowledge of the latter is needed, a graph can be \textit{polished} restricting the multiple sequence alignment problem only on the homologous sequences contained in each pancontig.

@@

After explaining the procedure we use for merging, we comment briefly on the effect of \textit{PanGraph}'s parameters on the resulting pangenome graphs.\\

@@

The choice of alignment kernel and the value of $\beta$ set instead a limit on the divergence of sequences that are merged in the same pancontig. The first intrinsic limit is set by the choice of alignment kernel and its sensitivity: minimap2 is significantly faster than mmseqs2, but it is in general unable to find matches between homologous sequences with more than 20\% divergence (with asm20 sensitivity).\\
On top of this intrinsic limit of the aligner, a more fine-tuned setting of the acceptable amount of divergence can be achieved by tuning the value of $\beta$. From the definition of the pseudo-energy (cf. eq.~(1) in the main text) it follows that matches between homologous sequence with divergence higher than $1/\beta$ will have positive pseudo-energy and will thus not result in a merge. If the molecular clock hypothesis holds, tuning this parameter is equivalent to setting an effective time limit beyond which sequences are not to be merged in the same alignment. This parameter can be tuned based on the diversity of the sample in study. It can for example be useful in resolving situations in which the hypothesis of a star-like phylogeny in a pancontig alignment is broken, such as to separate loosely related paralogs, as discussed in the next section.

@@

This paralog splitting can be done using two different sources of information: either sequence diversity or local context in which the duplication is inserted. Currently \textit{PanGraph} does not provide a specific command to perform paralog splitting. This is a feature that we plan on implementing in the future. However partial control of paralog split can currently be achieved in two ways.\\
The first one is based on sequence divergence. By tuning the value of the energy parameter $\beta$ the user can set a divergence limit. This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$ during graph construction.\\
The second method relies instead on context. The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag. When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context. These duplicated pancontigs are then split and absorbed in their context, thus simplifying the graph and loosing trace of the duplication.\\

@@

To test the accuracy of \textit{PanGraph} in reconstructing the pangenome graph of a set of isolates we generated artificial data using the procedure described in Section~\ref{sec:data_generation}. For each simulation we obtain both a set of genome sequences and the underlying real pangenome graph. This was done using standard value of the parameters, but varying the rate of HGT in the interval $h=[0.01,1]$ and the mutation rate $\mu=[0,0.01]$. For each $(h,\mu)$ pair 25 different sets of data were generated.\\
On each set of data we executed \textit{PanGraph} with values for the energy parameters $\alpha=0$ and $\beta=0$, and with three different alignment kernels:
\begin{itemize}
    \itemsep0pt
    \item \textit{minimap2} with option \textit{asm10},
    \item \textit{minimap2} with option \textit{asm20},
    \item \textit{mmseqs2},
\end{itemize}
thus obtaining three different pangenome graphs per dataset.\\
To link the mutation rate parameter $\mu$ with the average pairwise SNPs distance $\avg{d}$ between homologous segments in the population we evaluated the average pairwise distance for every pancontig in the pangenome graph with depth greater than one. For each graph we then perform the weighted average of these divergences, using as weight the length of the pancontig. Finally, for every pair of parameters $(\mu,h)$ we average these numbers over the 25 different simulations. Results are displayed in Fig.~\ref{fig:snps-suppl}.\\
For high values of $\mu$, the average pairwise divergence of pancontigs is not influenced by the rate of HGT $h$, but depends strongly on the choice of alignment kernel. With the \textit{asm20} option, \textit{minimap2} is able to correctly merge genomes with average pairwise divergence $\avg{d} \sim 7\%$ in our simulations, while \textit{mmseqs2} reaches $\avg{d} \sim 10\%$ at the expense of higher processing time. Notice that these values are lower than the threshold sensitivity of these aligners, which are expected to find matches up to respectively around 10\%, 20\% and 30\% sequence divergence. This is due to the fact that homologous sequences in our simulations are generated by an evolutionary process, and they can have a wide distribution of diversities and hierarchical population structure. Inability to merge sequences above a threshold divergence in a single pancontig is expected to result in multiple pancontigs containing sub-clades with higher similarity, and average diversity that is appreciably lower than the aligner threshold.\\
By performing a linear fit on the datapoints with $\mu \leq 0.002$ we are able to recover the conversion factor between the mutation rate and average pairwise divergence of pancontigs in our simulations: $\avg{d} \sim 20.8 \, \mu$.

@@

\begin{figure}[htb]
    \includegraphics[width=\textwidth]{figs_suppl/accuracy_comparison.pdf}
    \caption{{\bf Accuracy of PanGraph algorithm on synthetic data}. For each of our generated datasets we evaluate the average displacement of inferred pancontig breakpoints. The figure represents the cumulative distribution of average misplacements, stratified by the average pairwise divergence of sequences in pancontigs (see legend on the right). Each panel corresponds to a different alignment kernel used to infer the pangenome graph (see panel title). As divergence increases we observe a transition in the precision of breakpoint position estimation. At low divergence the position of most breakpoints is correctly estimated withing a few bps, while at high divergence most breakpoints can be misplaced by multiple hundreds of bps. The transition value for the divergence depends on the alignment kernel used, with \textit{mmseqs2} having the highest accuracy at the cost of higher computational time.}
    \label{fig:benchmark-accuracy-suppl}
\end{figure}

@@

We downloaded sets of annotated and complete chromosomes from RefSeq \cite{o2016reference} for 5 different bacterial species: \textit{Klebsiella pneumoniae} (KP), \textit{Helicobacter pylori} (HP), \textit{Prochlorococcus marinus} (PM), \textit{Mycobacterium tuberculosis} (MT) and \textit{Escherichia coli} (EC). In Table~I in the main text we report some summary statistics for these datasets. The number of isolates per species ranges from several hundreds for EC to 10 for PM.\\
These data has been chosen as part of a PanX \cite{ding2018panx} analysis dated March 2018.\footnote{
    The list of accession numbers is available on the repository: \url{https://github.com/neherlab/pangraph/blob/master/script/config/accession_numbers.csv}.
    We report for completeness that two of the isolates from the \textit{E.~coli} dataset from this analysis, namely \texttt{NC\_020518} and  \texttt{NC\_017663}, were subsequently removed from RefSeq. We verified that their removal from the dataset does not significantly impact core-genome and soft-core genome size, but decided to utilize the complete dataset as per original PanX analysis.
}
PanX is a pipeline for the analysis of bacterial pangenome. The pipeline takes as input a set of annotated genomes, and it clusters and aligns homologous genes. We used the PanX analysis to estimate the dataset diversity.\\
From this analysis we evaluated the average pairwise sequence divergence $\avg{d_\text{core}}$ of core genes, the average genome length $\avg{\Lgen}$, the total pangenome length $\Lpang$, the total core genome length $\Lcore$, and the total soft-core genome length $\Lsoftcore$, where soft-core genes are defined as genes being present in at least 95\% of the isolates. These quantities are reported in Table~I in the main text.\footnote{
    $\Lcore$ is the length of the alignment used to estimate sequence divergence $\avg{d_\text{core}}$. However this quantity is sensitive to annotation errors. This is particularly evident in the \textit{E.~coli} dataset, for which $\Lcore = 0.7$ Mbp is likely to be an underestimation of the real core genome size. $\Lsoftcore$ is instead less sensitive to annotation errors and usually considered a better core-genome size estimate. For reference, the \textit{E.~coli} core genome is usually considered to be around 1500-2500 genes in size, depending on the definition used \cite{park2019large,sutton2021pan,horsfield2023accurate}.
}\\
EC is the dataset with the highest number of isolates (307) and is also the one with the biggest pangenome. MT is the species with the smallest sequence divergence ($\dcore \sim 0.03\%$) and the biggest core genome fraction. The two datasets with highest sequence divergence are HP ($\dcore \sim 4\%$) and PM ($\dcore \sim 24\%$). For HP the average sequence divergence sits on the limit of what \textit{PanGraph} is able to accurately merge when using the \textit{minimap2} kernel (cf. Fig.~\ref{fig:benchmark-accuracy-suppl}), but is merged correctly by the \textit{mmseqs2} kernel. The sequence divergence of PM sits instead beyond the capabilities of all alignment kernels.

@@

We used genomes from the dataset described in the previous section to verify that marginalized graphs are compatible with graphs built directly from subset of strains. For each species we selected 50 different isolates (10 for PM) and built pangenome graphs using standard parameter values and the \textit{minimap2} kernel with \textit{asm20} option. For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{
    The list of selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/master/script/config/marginalize_pairs.csv}
}
For each pair we then built pairwise graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf. Fig.~6A, main text).\\
The comparison was performed by considering that each graph partitions a genome into pancontigs. Each pancontig can either be shared with the other member of the pair, or can be private to the isolate (cf. Fig.~6A, main text). For each isolate we consider the two partitions defined by the marginalized and pairwise graphs, and generate their intersection. Each block in this intersection can belong to two different categories, depending on whether the two partitions agree or disagree on whether the block is shared. Moreover, we can separate blocks on which we find agreement in two further categories: blocks that are shared and blocks that are private on both graphs.\\
For each pair of genomes and each type of block we evaluate the fraction of the genome that they cover and their average size. In Fig.~\ref{fig:marginalize-suppl} we display the distribution of these quantities across all pairs for the different species considered and the block types described above.\\
In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default pancontig length threshold of \textit{PanGraph} ($\Lthr =$100 bp).
As described in the previous section, the divergence of the HP dataset sits on the limit of what the \textit{minimap2} alignment kernel can successfully handle. As such, the results of merging different graphs together can depend on the order of merging, and pairwise and marginalized graphs can potentially show some disagreement. In this case the disagreement remains minor and only extends to few percents of the genome size, with disagreement blocks having size only slightly bigger than the 100 bp threshold of \textit{PanGraph}.
PM represents instead an example of a dataset with divergence much beyond the limit of the alignment kernel ($\dcore \sim 27 \%$). As a consequence, highly diverged homologous sequences are not merged, and the great majority of segments are private and have large size. Results also show a strong order-dependence, with some \textit{disagreement} blocks having size of 10s of kbp.
